---
title: "Consolidated Census Data"
---

{{< include ../../start-up.qmd >}}

```{r include=FALSE}
library(LSTbook)
People <- readr::read_csv("https://qr2-w25.github.io/homepage/projects/Census/Sunday.csv")
```

The *persons* spreadsheets from the Census project have been consolidated together. The data as of Sunday 9 March are available in CSV format [here](https://qr2-w25.github.io/homepage/projects/Census/Sunday.csv).



You can start to play with the data, perhaps looking for questions you want to attempt to use for your project. For example, here's a plot of education level versus age.

```{r warning=FALSE}
People |>
  point_plot(edu ~ age)
```

Questions:

1. What strong, familiar pattern do you see?
2. Can you spot some cases to investigate as outliers, perhaps pointing to incorrectly recorded or transcribed data?

## For easy reference

Variable names
```{r}
People |> names()
```


A few entries from each variable:

```{r}
People |> take_sample(n = 50) |> glimpse()
```

Keep in mind that the data are not well cleaned. And there's a lot of missing data. For instance, let's look at `wages` as transcribed by different people or groups. Here we look at the fraction (in percent) of missing data in several variables

```{r}
#| warning: false
People |> select(wages, age, edu, race, OWNER) |>
  summarize(
    wages_missing = round(100 * mean(is.na(wages))),
    age_missing  = round(100 * mean(is.na(age))),
    edu_missing  = round(100 * mean(is.na(edu))),
    race_missing = round(100 * mean(is.na(race))),
    .by = OWNER
  )
```

If you or your group are not listed in the above, it means `dkaplan@uaustin.org` doesn't yet have access to your data files.

## Exploring

Of course, you will want to ask your own questions ... 

```{webr-r}
#| autorun: true
#| echo: false
data_source <- url(
  "https://qr2-w25.github.io/homepage/projects/Census/Sunday.csv")
People <- readr::read_csv(data_source)
```

Start with making graphics of matters that interest you ...

```{webr-r}
People |> 
  point_plot( ..... )
```

Remember wrangling statements like `filter(sex == "F")` or `filter(wages > 0)`.

And then move on to modeling.

```{webr-r}


```


## Some empty R chunks

For your convenience.

```{webr-r}

```


```{webr-r}

```

```{webr-r}

```

`r devoirs_start_no_link()`

```{r echo=FALSE}
devoirs_end()
```
