---
title: "Kaplan's in-class notes for 2025-01-28"
date: 2025-01-28
---

```{r include=FALSE}
library(LSTbook)
```

## Go over quiz.

Question 3: What is a "unit of observation" in a data frame?

Question 4: Distinction between response and explanatory variable.

- It is the human modeler's choice. That choice depends on the goals motivating your analysis of data.
- There need not be an **causal** relationship between the explanatory variables and the response variable.
    - In *prediction*, often **association** is all that's needed, and causality, if any, might even go in the opposite direction.
    - In Block 5, we'll take on the issue of causality directly. For now, just understand that causality is a goal chiefly when we want to anticipate what change in output will result from a given change in input. 

Question 5. Make this plot ...

```{r echo=FALSE}
Penguins |> 
  point_plot(flipper ~ mass + island + sex, annot = "model")
```

Question 6: In what ways is this table **not** a data frame:

| brightness | distance | red shift | type
| -----------|----------|-----------|------
 | 3 | 2 parsec | 5% | M
 | 3 | 6 light-year | 0.07 | M
 | 2 | 8 | .09 | S
 | 1.3 | 19.2 | 10% | S
total | 9.3 | 35.2 | 2M 2S


A real-world example of this: [people in the DOD](www/rg2409.pdf) [Source](https://dwp.dmdc.osd.mil/dwp/app/dod-data-reports/workforce-reports)

- How to format this into a proper data frame. (Hint: Many ways.)
    - Start with: What is the *unit of observation*?
- Imagine what it would look like if the unit of observation were a single person.



Question 7:

- response is always on vertical axis
- the first explanatory variable is always on the x axis. 
- `annot = "violin"` sketches a nice picture of the distribution.

Show what happens as $n$ gets bigger.

```{webr-r}
Juice_prices <- datasim_make(
  juice <- categorical(n, "orange", "apple"),
  price <- round(rnorm(n, mean = 1.75, sd=0.50), 2) +
    cat2value(juice, orange = 0.47, apple = 0 )
)
Juice_prices |>
  take_sample(n=20) |>
  point_plot(price ~ juice, annot = "violin")
```

## Census project

[1940 Population schedule form and codebook](https://www.archives.gov/research/census/1940/research/census/1940/general-info#form)

[Documents for Travis County, Texas](https://catalog.archives.gov/search-within/57229197)

[A population schedule](www/m-t0627-04148-00916.jpg) from Austin

Optaining enumeration districts: https://stevemorse.org/census/

https://stevemorse.org/census/schedules.php

4148. Travis; Travis (Austin) (227-1 to 227-7, 227-8 to 227-26 )
4149. Travis (Austin) (227-27 to 227-47 )
4150. Travis (Austin) (227-48 to 227-71 )



## Spiegelhalter Chaps 5 and 6

