---
title: "Kaplan's in-class notes for 2025-01-28"
date: 2025-01-28
---

```{r include=FALSE}
library(LSTbook)
```

## Go over quiz.

Question 3: What is a "unit of observation" in a data frame?

Question 4: Distinction between response and explanatory variable.

- It is the human modeler's choice. That choice depends on the goals motivating your analysis of data.
- There need not be an **causal** relationship between the explanatory variables and the response variable.
    - In *prediction*, often **association** is all that's needed, and causality, if any, might even go in the opposite direction.
    - In Block 5, we'll take on the issue of causality directly. For now, just understand that causality is a goal chiefly when we want to anticipate what change in output will result from a given change in input. 

Question 5. Make this plot ...

```{r echo=FALSE}
Penguins |> 
  point_plot(flipper ~ mass + island + sex, annot = "model")
```

Question 6: In what ways is this table **not** a data frame:

| brightness | distance | red shift | type
| -----------|----------|-----------|------
 | 3 | 2 parsec | 5% | M
 | 3 | 6 light-year | 0.07 | M
 | 2 | 8 | .09 | S
 | 1.3 | 19.2 | 10% | S
total | 9.3 | 35.2 | 2M 2S


A real-world example of this: [people in the DOD](www/rg2409.pdf) [Source](https://dwp.dmdc.osd.mil/dwp/app/dod-data-reports/workforce-reports)

- How to format this into a proper data frame. (Hint: Many ways.)
    - Start with: What is the *unit of observation*?
- Imagine what it would look like if the unit of observation were a single person.



Question 7:

- response is always on vertical axis
- the first explanatory variable is always on the x axis. 
- `annot = "violin"` sketches a nice picture of the distribution.

Show what happens as $n$ gets bigger.

```{webr-r}
Juice_prices <- datasim_make(
  juice <- categorical(n, "orange", "apple"),
  price <- round(rnorm(n, mean = 1.75, sd=0.50), 2) +
    cat2value(juice, orange = 0.47, apple = 0 )
)
Juice_prices |>
  take_sample(n=20) |>
  point_plot(price ~ juice, annot = "violin")
```

## Census project

[1940 Population schedule form and codebook](https://www.archives.gov/research/census/1940/research/census/1940/general-info#form)

[Documents for Travis County, Texas](https://catalog.archives.gov/search-within/57229197)

[A population schedule](www/m-t0627-04148-00916.jpg) from Austin

Optaining enumeration districts: https://stevemorse.org/census/

https://stevemorse.org/census/schedules.php

4148. Travis; Travis (Austin) (227-1 to 227-7, 227-8 to 227-26 )
4149. Travis (Austin) (227-27 to 227-47 )
4150. Travis (Austin) (227-48 to 227-71 )



## Spiegelhalter Chaps 5 and 6

### Speed cameras

1. No, he does not think speed cameras are effective. He thinks that accidents
*would have returned back to normal anyways*, without intervention. Since a
high accident rate is abnormal, "regression to the mean" will lead the rate
to return to normal naturally and cannot be attributed to the installment of
speed cameras. []

2. Speigelhalter says that speed cameras are **less effective than they seem** because of regression to the mean. Regression to the mean describes the fact that
when something is unusual (such as higher than mean crashes) it is *usually
just because of luck*, therefore the luck is bound to run out and the expected
result is that it return closer to mean levels. If speed cameras are put in
in response to unusually high crash rates, and then there are fewer crashes
it’s *probably not due to the camera but just because the bad luck has run
out*.

[The question is how to separate out the effects of "regression to the mean" from the actual effect of the intervention. HOW CAN WE DO THIS?]

3.Likely not because of their tendency to be put in places after wrecks. The
Camera was placed after an extreme event and it reports normal traffic which
makes it seem like the rode got safer when in reality the rode was just
returning to normal after the wreck, a regression to the mean.

4. Speed cameras are only effect about a third of the time, most of the time when
speed cameras are put in place it is to reduce traffic violations in that
area, but regression to the mean suggest that the speed cameras weren't the
reason at all. Regression to the mean is when things gradually shift towards
the mean, regardless of outside factors."
They can be effective, but it is important to perform a regression to determine
what variable is causing the change.

5. According to Spiegelhalter, *speed cameras are not practical all the time*. A
history of bad luck in accidents resulting in the implementation of the speed
camera does not mean the lower rate of accidents now is due to the camera.
It could be due to the accident rates dropping back to the average rate when
they occur. Spiegelhalter points out that we give intervention the credit
when it could be a reversion to the average (regression to the mean)."

6. Yes, speed cameras are two-thirds effective, this was determined by modeling a
regression that shows a tendency for them to work.

7. There is evidence that speed cameras can improve safety, but Spiegelhalter
acknowledges that there is evidence suggesting that the "effectiveness" of
speed cameras is often overstated due to statistical principles. Regarding
“regression to the mean”, speed cameras are installed at locations that have
recently experienced a high number of accidents but this decision is often
an assumption. As a result, the reduction in accidents after installing
speed cameras might not be entirely due to the cameras but instead due to
regression to the mean.

8. possibly, though if installed because of a recent high accident rate, may not be the cause of the reduction (or have their effect overestimated) as the accident rate would have likely regressed to the mean regardless of the intervention

9. Spiegelhalter acknowledges that speed cameras are often associated with reducing accidents at locations where they are installed. However, he points out that this observed decrease can partly be explained by regression to the mean. RTTM is important as places where cameras are located usually have higher than average accident rates and would have regressed back to the average rate without intervention.

### Placebo effect

> The placebo effect is a perceived and then actual benefit from people in the control group who think they are receiving treatment. This improvement can be seen as "regression to the mean" because these patients may have just had conditions that resolved themselves anyways.

### Logistic regression

It makes a difference where you enter an answer!

> Subjects are often selected for their extreme symptoms, which may naturally regress to the mean due to random variation and not the treatment. regression to the mean and the psychosomatic placebo effect could both confound the effect of treatment and make the illusion of efficacy.

Go through 5-before questions and 6-follow-up.

## Computing

Contrast prediction with regression: Same model training, but different goal: to understand what the connections are in a system.

- Different criteria apply to the selection of explanatory variables.
- We are interested in details of the model, not just the prediction. But the form will look much like he prediction interval for quantitative response variables.

```{webr-r}
Galton |>
  point_plot(height ~ mother + 1 + sex, annot = "model", interval = "prediction")
```

- Show also "confidence band"

```{webr-r}
Hmod1 <- Galton |>
  model_train(height ~ mother + father + sex) 
Hmod1 |> conf_interval()
```
